\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-01-28 submission/manuscript.tex
Rerender

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (44). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (44). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al.~(available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (45), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (43, 46, 47);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

\leavevmode\hypertarget{ref-dormann_collinearity:_2013}{}%
37. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
of methods to deal with it and a simulation study evaluating their
performance. Ecography \textbf{36}:27--46.
doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.

\leavevmode\hypertarget{ref-li_accurate_2020}{}%
38. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
Accurate data-driven prediction does not mean high reproducibility. Nat
Mach Intell \textbf{2}:13--15.
doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.

\leavevmode\hypertarget{ref-sze_fecal_2019}{}%
39. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
acids are not predictive of colonic tumor status and cannot be predicted
based on bacterial community structure. mBio \textbf{10}:e01454--19.
doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.

\leavevmode\hypertarget{ref-kocheturov_massive_2019}{}%
40. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
2019. Massive datasets and machine learning for computational
biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.

\leavevmode\hypertarget{ref-kim_improved_2018}{}%
41. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
method for prediction of cancer prognosis by network learning. Genes
\textbf{9}:478.
doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.

\leavevmode\hypertarget{ref-wiens_no_2019}{}%
42. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
2019. Do no harm: A roadmap for responsible machine learning for health
care. Nat Med \textbf{25}:1337--1340.
doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.

\leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
43. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\leavevmode\hypertarget{ref-redwood_stool_2016}{}%
44. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
for screening detection of colorectal neoplasia in alaska native people.
Mayo Clin Proc \textbf{91}:61--70.
doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.

\leavevmode\hypertarget{ref-sze_leveraging_2018}{}%
45. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
46. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\leavevmode\hypertarget{ref-rognes_vsearch_2016}{}%
47. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\leavevmode\hypertarget{ref-li_hyperband:_2016}{}%
48. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
bandit-based approach to hyperparameter optimization. arXiv:160306560
{[}cs, stat{]}.

>>>>

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (5). Stool
samples were provided by adults who were undergoing a scheduled
screening or surveillance colonoscopy. Participants were recruited from
Toronto (ON, Canada), Boston (MA, USA), Houston (TX, USA), and Ann Arbor
(MI, USA). Patients' colonic health was visually assessed by colonoscopy
with bowel preparation and tissue histopathology of all resected
lesions. We assigned patients into two classes: those with healthy
colons and those with screen relevant neoplasias (SRNs). The healthy
class included patients with healthy colons or non-advanced adenomas,
whereas the SRN class included patients with advanced adenomas or
carcinomas (46). Patients with an adenoma greater than 1 cm, more than
three adenomas of any size, or an adenoma with villous histology were
classified as having advanced adenomas (46). There were 172 patients
with normal colonoscopies, 198 with adenomas, and 120 with carcinomas.
Of the 198 adenomas, 109 were identified as advanced adenomas. Together
261 patients were classified as healthy and 229 patients were classified
as having an SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al.~(available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (47), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (45, 48, 49);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

\leavevmode\hypertarget{ref-breiman_statistical_2001}{}%
37. \textbf{Breiman L}. 2001. Statistical modeling: The two cultures
(with comments and a rejoinder by the author). Statist Sci
\textbf{16}:199--231.
doi:\href{https://doi.org/10.1214/ss/1009213726}{10.1214/ss/1009213726}.

\leavevmode\hypertarget{ref-strobl_bias_2007}{}%
38. \textbf{Strobl C}, \textbf{Boulesteix A-L}, \textbf{Zeileis A},
\textbf{Hothorn T}. 2007. Bias in random forest variable importance
measures: Illustrations, sources and a solution. BMC Bioinformatics
\textbf{8}:25.
doi:\href{https://doi.org/10.1186/1471-2105-8-25}{10.1186/1471-2105-8-25}.

\leavevmode\hypertarget{ref-dormann_collinearity:_2013}{}%
39. \textbf{Dormann CF}, \textbf{Elith J}, \textbf{Bacher S},
\textbf{Buchmann C}, \textbf{Carl G}, \textbf{Carré G}, \textbf{Marquéz
JRG}, \textbf{Gruber B}, \textbf{Lafourcade B}, \textbf{Leitão PJ},
\textbf{Münkemüller T}, \textbf{McClean C}, \textbf{Osborne PE},
\textbf{Reineking B}, \textbf{Schröder B}, \textbf{Skidmore AK},
\textbf{Zurell D}, \textbf{Lautenbach S}. 2013. Collinearity: A review
of methods to deal with it and a simulation study evaluating their
performance. Ecography \textbf{36}:27--46.
doi:\href{https://doi.org/10.1111/j.1600-0587.2012.07348.x}{10.1111/j.1600-0587.2012.07348.x}.

\leavevmode\hypertarget{ref-li_accurate_2020}{}%
40. \textbf{Li J}, \textbf{Liu L}, \textbf{Le TD}, \textbf{Liu J}. 2020.
Accurate data-driven prediction does not mean high reproducibility. Nat
Mach Intell \textbf{2}:13--15.
doi:\href{https://doi.org/10.1038/s42256-019-0140-2}{10.1038/s42256-019-0140-2}.

\leavevmode\hypertarget{ref-sze_fecal_2019}{}%
41. \textbf{Sze MA}, \textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2019. Fecal short-chain fatty
acids are not predictive of colonic tumor status and cannot be predicted
based on bacterial community structure. mBio \textbf{10}:e01454--19.
doi:\href{https://doi.org/10.1128/mBio.01454-19}{10.1128/mBio.01454-19}.

\leavevmode\hypertarget{ref-kocheturov_massive_2019}{}%
42. \textbf{Kocheturov A}, \textbf{Pardalos PM}, \textbf{Karakitsiou A}.
2019. Massive datasets and machine learning for computational
biomedicine: Trends and challenges. Ann Oper Res \textbf{276}:5--34.
doi:\href{https://doi.org/10.1007/s10479-018-2891-2}{10.1007/s10479-018-2891-2}.

\leavevmode\hypertarget{ref-kim_improved_2018}{}%
43. \textbf{Kim M}, \textbf{Oh I}, \textbf{Ahn J}. 2018. An improved
method for prediction of cancer prognosis by network learning. Genes
\textbf{9}:478.
doi:\href{https://doi.org/10.3390/genes9100478}{10.3390/genes9100478}.

\leavevmode\hypertarget{ref-wiens_no_2019}{}%
44. \textbf{Wiens J}, \textbf{Saria S}, \textbf{Sendak M},
\textbf{Ghassemi M}, \textbf{Liu VX}, \textbf{Doshi-Velez F},
\textbf{Jung K}, \textbf{Heller K}, \textbf{Kale D}, \textbf{Saeed M},
\textbf{Ossorio PN}, \textbf{Thadaney-Israni S}, \textbf{Goldenberg A}.
2019. Do no harm: A roadmap for responsible machine learning for health
care. Nat Med \textbf{25}:1337--1340.
doi:\href{https://doi.org/10.1038/s41591-019-0548-6}{10.1038/s41591-019-0548-6}.

\leavevmode\hypertarget{ref-westcott_opticlust_2017}{}%
45. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
Improved Method for Assigning Amplicon-Based Sequence Data to
Operational Taxonomic Units. mSphere \textbf{2}.
doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.

\leavevmode\hypertarget{ref-redwood_stool_2016}{}%
46. \textbf{Redwood DG}, \textbf{Asay ED}, \textbf{Blake ID},
\textbf{Sacco PE}, \textbf{Christensen CM}, \textbf{Sacco FD},
\textbf{Tiesinga JJ}, \textbf{Devens ME}, \textbf{Alberts SR},
\textbf{Mahoney DW}, \textbf{Yab TC}, \textbf{Foote PH}, \textbf{Smyrk
TC}, \textbf{Provost EM}, \textbf{Ahlquist DA}. 2016. Stool DNA testing
for screening detection of colorectal neoplasia in alaska native people.
Mayo Clin Proc \textbf{91}:61--70.
doi:\href{https://doi.org/10.1016/j.mayocp.2015.10.008}{10.1016/j.mayocp.2015.10.008}.

\leavevmode\hypertarget{ref-sze_leveraging_2018}{}%
47. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\leavevmode\hypertarget{ref-schloss_introducing_2009}{}%
48. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.

\leavevmode\hypertarget{ref-rognes_vsearch_2016}{}%
49. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
source tool for metagenomics. PeerJ \textbf{4}:e2584.
doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.

\leavevmode\hypertarget{ref-li_hyperband:_2016}{}%
50. \textbf{Li L}, \textbf{Jamieson K}, \textbf{DeSalvo G},
\textbf{Rostamizadeh A}, \textbf{Talwalkar A}. 2016. Hyperband: A novel
bandit-based approach to hyperparameter optimization. arXiv:160306560
{[}cs, stat{]}.
